---
title: "Lab 8 - Ecological Niche Modeling"
author: "ECOL 620 - Applications in Landscape Ecology"
date: "Spring 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lab, we explore methodologies for constructing species distribution models (SDM), including envelope models, GLMs, GAMs, and random forest This is an extensive area of study, and given our limited time to cover it, this lab is only intended as an introduction to the potential techniques. Think of this as a sampling platter of SDMs. We will use Varied Thrush surveys from northern Idaho and western Montana to build candidate distribution models.

Read “Fletcher_Chapter7.pdf” and open “lab7_practice.Rmd” in R. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab7_firstname_lastname.html (ex. lab7_kyle_horton.html).  Note, we will not grade labs in any other format.



```{r}
#load packages
library(raster)           #for raster covariate data; 
library(reshape2)         #for re-formatting data
library(mgcv)             #for gams
library(dismo)            #for SDMs
library(randomForest)     #for random forest SDMs
library(MuMIn)            #for model selection
library(PresenceAbsence)  #for model evaluation
library(ecospat)         #for model evaluation
```

#library(devtools) #this might be necessary because the newest version (1.3-3) generates an 
#issue with the percRank function 
#install_version("dismo", version = "1.1-4", repos = "http://cran.us.r-project.org")

## Preparing the data


### subsetting point data
```{r}
vath.data <- read.csv(file="/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week11/Lab8/data_for_lab8/vath_2004.csv", header=TRUE)
vath.val <- read.csv(file="/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week11/Lab8/data_for_lab8/vath_VALIDATION.csv", header=TRUE)
head(vath.data)
table(vath.data$VATH)
```


```{r}
#subset to presence-only / absence-only
vath.pres <- vath.data[vath.data$VATH==1,]
vath.abs <- vath.data[vath.data$VATH==0,]
vath.pres.xy <- as.matrix(vath.pres[,c("EASTING","NORTHING")])
vath.abs.xy <- as.matrix(vath.abs[,c("EASTING","NORTHING")])
```



```{r}
#validation data
vath.val.pres <- as.matrix(vath.val[vath.val$VATH==1, c("EASTING","NORTHING")])
vath.val.abs <- as.matrix(vath.val[vath.val$VATH==0, c("EASTING","NORTHING")])
vath.val.xy <- as.matrix(vath.val[,c("EASTING","NORTHING")])
```



### viewing GIS data

```{r}
#covariate maps
elev <- raster("/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week11/Lab8/data_for_lab8/elev.gri")                 #elevation layer
canopy <- raster("/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week11/Lab8/data_for_lab8/cc2.gri")                #linear gradient in canopy cover taken from PCA
mesic <- raster("/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week11/Lab8/data_for_lab8/mesic.gri")               #presence of mesic forest
precip <- raster("/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week11/Lab8/data_for_lab8/precip.gri")             #mean precip (cm)
```


```{r}
#check maps
compareRaster(elev, canopy)
# compareRaster(elev, mesic) # note, if you uncomment this line, it should produce an error, which is an issue
# compareRaster(elev, precip) # note, if you uncomment this line, it should produce an error, which is an issue
```
```{r}
#resample to align layers
mesic <- resample(x=mesic, y=elev, "ngb")            #nearest neighbor (categorical)
precip <- resample(x=precip, y=elev, "bilinear")     #for continuous data
```


```{r}
#crop to same extent
mesic <- mask(mesic, elev)
precip <- mask(precip, elev)
```

```{r}
#check maps
compareRaster(elev,precip, mesic)
```



```{r}
#make 1 km wet forest
fw.1km <- focalWeight(mesic, 1000, 'circle')           #buffer in CRS units
mesic1km <- focal(mesic, w=fw.1km, fun="sum", na.rm=T)
```


```{r}
#create raster stack
layers <- stack(canopy, elev, mesic, mesic1km, precip)
names(layers) <- c("canopy", "elev", "mesic", "mesic1km", "precip")
```


```{r}
#plot stack and correlations among covariates
pairs(layers, maxpixels=1000)                          #maxpixels sets upper limit on sampling raster
plot(layers)
```

```{r}
#drop correlated layer (mesic)
layers <- dropLayer(layers, 3)
```


```{r}
#Generate availability/background points using dismo
back.xy <- randomPoints(layers, p=vath.pres.xy, n=2000)
```

```{r}
#inspect
head(back.xy)
```


```{r}
#re-name columns
colnames(back.xy) <- c("EASTING","NORTHING")
```

```{r}
#plot
plot(elev)
points(back.xy)
```

```{r}
#extract GIS data
pres.cov <- extract(layers, vath.pres.xy)          #extracts values from layers at pres locations
back.cov <- extract(layers, back.xy)               #extracts values from layers at random locations
val.cov <- extract(layers, vath.val.xy)            #extracts values from layers at validation locations
```

```{r}
#link data
pres.cov <- data.frame(vath.pres.xy, pres.cov, pres=1)
back.cov <- data.frame(back.xy, back.cov, pres=0)
val.cov <- data.frame(vath.val, val.cov)
```

```{r}
#remove any potential NAs
pres.cov <- pres.cov[complete.cases(pres.cov),]
back.cov <- back.cov[complete.cases(back.cov),]
val.cov <- val.cov[complete.cases(val.cov),]
```


```{r}
#bind presence and background points together
all.cov <- rbind(pres.cov, back.cov)
```


```{r}
#inspect
head(all.cov)
```



# #######################################################################################
# Envelopes

```{r}
#fit model
bioclim.vath <- bioclim(layers, vath.pres.xy)
```


```{r}
#inspect
summary(bioclim.vath)
names(layers)
```


```{r}
#plot
plot(bioclim.vath, a=1, b=2, p=0.95)        #elev-canopy plot 85% quantile bounding box
plot(bioclim.vath, a=1, b=2, p=0.90)        #elev-canopy plot 95% quantile bounding box
plot(bioclim.vath, a=1, b=4, p=0.95)        #elev-precip plot
```

```{r}
#mapping
bioclim.map <- predict(layers, bioclim.vath)

```

```{r}
#plot
plot(bioclim.map, axes=F, box=F, main="bioclim")
```



# ######################################################################################
# GLMs and GAMs

## GLMs

```{r}
glm.vath <- glm(pres~canopy+elev+I(elev^2)+mesic1km+precip, family=binomial(link=logit), data=all.cov)
```


```{r}
#inspect
summary(glm.vath)
```


```{r}
#mapping
glm.map <- predict(layers, glm.vath, type="response")
```


```{r}
#plot
plot(glm.map, axes=F, box=F, main="GLM")
```


# GAMs

```{r}
#GAM (default settings with optimal knots determined by generalized cross validation)
gam.vath <- gam(pres~s(canopy)+s(elev)+s(mesic1km)+s(precip), family=binomial(link=logit), method="ML", data=all.cov)
```

```{r}
#inspect
summary(gam.vath)
```

```{r}
#plot relationships
plot(gam.vath, shade=T)
```


```{r}
#Manually alter the number of knots
gam.vath.knot3 <- gam(pres~s(canopy,k=3)+s(elev,k=3)+s(mesic1km,k=3)+s(precip,k=3), family=binomial(link=logit), method="ML", data=all.cov)
gam.vath.knot6 <- gam(pres~s(canopy,k=6)+s(elev,k=6)+s(mesic1km,k=6)+s(precip,k=6), family=binomial(link=logit), method="ML", data=all.cov)
summary(gam.vath.knot3)
summary(gam.vath.knot6)
```

```{r}
#plot relationships and compare
plot(gam.vath.knot3, shade=T)
plot(gam.vath.knot6, shade=T)
```

```{r}
#Consider interactions among splines with tensors (this is slow; ~ 6min)
gam.vath.tensor <- gam(pres~te(canopy,elev,precip,mesic1km), family=binomial(link=logit), method="ML", data=all.cov)

#plot
plot(gam.vath.tensor, shade=T)
```


```{r}
#Change the smoothing function
gam.vath.cr <- gam(pres~s(canopy, bs="cr")+s(elev, bs="cr")+s(mesic1km, bs="cr")+s(precip, bs="cr"), family=binomial(link=logit), method="ML", data=all.cov)

#plot
plot(gam.vath.cr, shade=T)
```


```{r}
#evaluation of gam tuning (with evaluate function in dismo)
eval.gam <- evaluate(p=vath.val.pres, a= vath.val.abs, gam.vath, layers)
eval.gam3 <- evaluate(p=vath.val.pres, a= vath.val.abs, gam.vath.knot3, layers)
eval.gamte <- evaluate(p=vath.val.pres, a= vath.val.abs, gam.vath.tensor, layers)
eval.gamcr <- evaluate(p=vath.val.pres, a= vath.val.abs, gam.vath.cr, layers)
```


```{r}
#inspect tuning
eval.gamcr
```


```{r}
#evaluation with AIC
round(AIC(gam.vath, gam.vath.knot3, gam.vath.knot6, gam.vath.tensor, gam.vath.cr), 1)

```

```{r}
#mapping
gam.map <- predict(layers, gam.vath.knot3, type="response")
```


```{r}
#plot
plot(gam.map, axes=F, box=F, main="GAM")
```


# #######################################################################################
# Random Forests

```{r}
#random forest model (default)
rf.vath <- randomForest(as.factor(pres) ~ canopy+elev+mesic1km+precip, na.action=na.omit, data=all.cov)
```


```{r}
#tuning model
rf.vath.tune <- tuneRF(y=as.factor(all.cov$pres), x = all.cov[,c(3:6)], stepFactor=0.5, ntreeTry=500)
```

```{r}
#update rf model with mtry=1 based on tuning
rf.vath <- randomForest(as.factor(pres) ~ canopy+elev+mesic1km+precip, mtry=1, ntree=500, na.action=na.omit, data=all.cov)
```


```{r}
#variable importance plot
varImpPlot(rf.vath)
```


```{r}
#mapping
rf.map <- predict(layers, rf.vath, type="prob",index=2)

#plot
plot(rf.map, axes=F, box=F, main="RF")
```



# ################################################################################################
# Interpreting environmental relationships

```{r}
#median of each variable
elev.median <- median(back.cov$elev, na.rm=T)
canopy.median <- median(back.cov$canopy, na.rm=T)
precip.median <- median(back.cov$precip, na.rm=T)
mesic1km.median <- median(back.cov$mesic1km, na.rm=T)

#range
elev.range <- seq(min(back.cov$elev, na.rm=T), max(back.cov$elev, na.rm=T), length=100)
canopy.range <- seq(min(back.cov$canopy, na.rm=T), max(back.cov$canopy, na.rm=T), length=100)
```

```{r}
#Data frame of new data
elev.partial.data <- data.frame(expand.grid(elev=elev.range, canopy=canopy.median, precip=precip.median, mesic1km=mesic1km.median))
canopy.partial.data <- data.frame(expand.grid(elev=elev.median, canopy=canopy.range, precip=precip.median, mesic1km=mesic1km.median))

```


```{r}
#Predict onto new data
bio.pred.elev <- predict(bioclim.vath, elev.partial.data)
bio.pred.canopy <- predict(bioclim.vath, canopy.partial.data)

glm.pred.elev <- predict(glm.vath, elev.partial.data,type="response")
glm.pred.canopy <- predict(glm.vath, canopy.partial.data,type="response")

gam.pred.elev <- predict(gam.vath, elev.partial.data,type="response")
gam.pred.canopy <- predict(gam.vath, canopy.partial.data,type="response")

rf.pred.elev <- predict(rf.vath, elev.partial.data, type="prob")
rf.pred.canopy <- predict(rf.vath, canopy.partial.data, type="prob")
rf.pred.elev <- rf.pred.elev[,2]
rf.pred.canopy <- rf.pred.canopy[,2]
```


```{r}
#max.pred.elev <- predict(max.vath, elev.partial.data)
#max.pred.canopy <- predict(max.vath, canopy.partial.data)

#Data frame for plots
part.elev.df <- data.frame(elevation=elev.range,
                       bioclim=bio.pred.elev, glm=glm.pred.elev,gam=gam.pred.elev,
                       rf=rf.pred.elev
                       #,max=max.pred.elev
                       )
part.canopy.df <- data.frame(canopy=canopy.range,
                       bioclim=bio.pred.canopy, glm=glm.pred.canopy,gam=gam.pred.canopy,
                       rf=rf.pred.canopy
                       #,max=max.pred.canopy
                       )
```


```{r}

#plot elevation
plot(part.elev.df$elevation, part.elev.df$bioclim, type='l', xlab="Elevation", ylab="Response", ylim=c(0,0.6))
lines(part.elev.df$elevation, part.elev.df$glm, type='l',col="red")
lines(part.elev.df$elevation, part.elev.df$gam, type='l',col="orange")
lines(part.elev.df$elevation, part.elev.df$rf, type='l',col="blue")
#lines(part.elev.df$elevation, part.elev.df$max, type='l',col="purple")
```

```{r}
#plot canopy
plot(part.canopy.df$canopy, part.canopy.df$bioclim, type='l', xlab="canopy", ylab="Response", ylim=c(0,0.7))
lines(part.canopy.df$canopy, part.canopy.df$glm, type='l',col="red")
lines(part.canopy.df$canopy, part.canopy.df$gam, type='l',col="orange")
lines(part.canopy.df$canopy, part.canopy.df$rf, type='l',col="blue")
#lines(part.canopy.df$canopy, part.canopy.df$max, type='l',col="purple")
```


# ##################################################################################################
# Model evaluation


#to use PresenceAbsence Package:
#data frame format:
#column 1: siteID; column 2: validation 0/1; column 3-N: model predictions (column 3 = model 1)


# evaluate based on prospective sampling

```{r}
#predictions for validation
val.cov.pred <- val.cov[,cbind("canopy", "elev", "mesic1km", "precip")]
bio.val <- predict(bioclim.vath, val.cov.pred)
glm.val <- predict(glm.vath, val.cov.pred, type="response")
gam.val <- predict(gam.vath, val.cov.pred, type="response")
rf.val <- predict(rf.vath, val.cov.pred, type="prob")
rf.val <- rf.val[,2]
#max.val <- predict(max.vath, val.cov.pred)
```


```{r}

#PresenceAbsence data frame
val.data <- data.frame(siteID=1:nrow(vath.val), obs=vath.val$VATH,
                      bio=bio.val, glm=glm.val, gam=gam.val, rf=rf.val
                      #, max=max.val
                      )

```

```{r}
#correlation among model predictions
round(cor(val.data[,c("bio","glm","gam","rf"
                      #,"max"
                      )], method="spearman"),2)
```

```{r}
#data frame to store summary statistics
summary.eval <- data.frame(matrix(nrow=0, ncol=9))
names(summary.eval) <- c("model", "auc", "corr", "ll", "threshold", "sens", "spec", "tss", "kappa")
```


```{r}

#sensitivity = true positive rate
#specificity = true negative rate
#tss = true skill statistic

nmodels <- ncol(val.data)-2

for(i in 1:nmodels){

  #calculate summary statistics
  auc.i <- auc(val.data, which.model=i)
  kappa.opt <- optimal.thresholds(val.data, which.model=i, opt.methods=3)
  sens.i <- sensitivity(cmx(val.data, which.model=i,threshold = kappa.opt[[2]]))
  spec.i <- specificity(cmx(val.data, which.model=i,threshold = kappa.opt[[2]]))
  tss.i<- sens.i$sensitivity +spec.i$specificity - 1
  kappa.i <- Kappa(cmx(val.data, which.model=i,threshold = kappa.opt[[2]]))
  corr.i <- cor.test(val.data[,2], val.data[,i+2])$estimate
  ll.i <- sum(log(val.data[,i+2]*val.data[,2] + (1-val.data[,i+2])*(1-val.data[,2])))
  ll.i <- ifelse(ll.i=="-Inf", sum(log(val.data[,i+2]+0.001)*val.data[,2] + log((1-val.data[,i+2]))*(1-val.data[,2])), ll.i)

  #summarize
  summary.i <- c(i,auc.i$AUC, corr.i, ll.i,kappa.opt[[2]], sens.i$sensitivity, spec.i$specificity, tss.i, kappa.i[[1]])
  summary.eval <- rbind(summary.eval, summary.i)
}
names(summary.eval) <- c("model", "auc", "corr", "ll", "threshold", "sens", "spec", "tss", "kappa")
```

```{r}
#inspect
summary.eval
```


```{r}
#add model names
summary.eval$model <- c("bio", "glm", "gam", "rf"
                        #, "max"
                        )
```



